On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-21 manuscripts/PNAS/chazthm_pnas.tex
sml edits fixed rozell ref

More formally, let $\mathbf{A} \in \mathbb R^{n \times m}$ be a matrix with columns $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our combined worst-case uncertainty in  measuring $\mathbf{A}\mathbf{x}_i$.
The first mathematical problem addressed here is the following.

@@

We generalize the  lower bound to a domain-restricted \emph{union of subspaces} model \cite{vidal2005generalized} derived from a hypergraph $\mathcal{H}$. Letting $\mathbf{M}_J$ denote the submatrix formed by the columns of $\mathbf{M}$ indexed by $J \subseteq [m]$ (with $\mathbf{M}_\emptyset := \mathbf{0}$), we define: 
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \inf \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \inf \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \alpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \frac{1}{\sqrt{k}} \inf \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^{|S|} \right\},
\end{align} 
%
where we write $L_{k}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{We note that $1 - \sqrt{k} L_k(\mathbf{M})$ is known as the asymmetric lower restricted isometry constant for matrices $\mathbf{M}$ with unit $\ell_2$-norm columns \cite{Blanchard2011}.} Clearly, $L_{2k}(\mathbf{M}) > 0$ for $\mathbf{M}$ satisfying \eqref{SparkCondition} and $L_{k'}(\mathbf{M}) \geq L_k(\mathbf{M})$ whenever $k' \leq k$. Note also that $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$ if $\mathcal{H} \subseteq {[m] \choose k}$ has $\cup \mathcal{H} = [m]$.

A vector $\mathbf{x}$ is said to be \emph{supported} in $S \subseteq [m]$ when $\mathbf{x} \in \text{\rm span}\{\mathbf{e}_j: j\in S\}$, where $\mathbf{e}_j$ are the standard column basis in $\mathbb R^m$. 
%Denote by $\mathbf{x}^J$ the subvector formed from the entries of $\mathbf{x}$ indexed by $J$. 
A set of $k$-sparse vectors is said to be in \emph{general linear position} when any $k$ of them are linearly independent. The following is a precise statement of our main result.  We leave the quantity $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, \mathcal{H})$ undefined until Eq.~\eqref{Cdef1}.

@@

If $\bar m < mr / (r-1)$ then, letting $p := \bar m - r( \bar m - m)$, we have $|T_1| = \bar m (r - 1) + p$ for positive $p$. By the pigeonhole principle again, there must then be at least $r$ elements of $T_1$ sharing the same first index. By \eqref{cond}, the intersection of the set $\mathcal{G}_1$ consisting of their second indices is nonempty; hence, $|\cap \mathcal{G}_1| = 1$ by $r$-regularity and the SIP. Since $\cap \pi(\mathcal{G}_1)$ is nonempty, applying \eqref{cond} again gives $\cap \pi(\mathcal{G}_1) = \{i_1\}$ for some $i_1 \in [m]$. If $p=1$, then we are done. Otherwise, define $T_2 := T_1 \setminus \{(i,S) \in T_1: i = i_1\}$, which contains $|T_2| = |T_1| - r = (\bar m - 1)(r-1) + (p-1)$ ordered pairs having $\bar m - 1$ distinct first indices. Repeating the above arguments produces a (necessarily) distinct index $i_2$. Iterating the arguments $p$ times yields the set of singletons \mbox{$J = \{i_1, \ldots, i_p\} \subseteq [m]$}.
\end{proof}

>>>>

More formally, let $\mathbf{A} \in \mathbb R^{n \times m}$ be a matrix with columns $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our combined worst-case uncertainty in  measuring $\mathbf{A}\mathbf{x}_i$.
The first mathematical problem we address is the following.

@@

We generalize the  lower bound to a domain-restricted \emph{union of subspaces} model \cite{vidal2005generalized} derived from a hypergraph $\mathcal{H}$. Letting $\mathbf{M}_J$ denote the submatrix formed by the columns of $\mathbf{M}$ indexed by $J \subseteq [m]$ (with $\mathbf{M}_\emptyset := \mathbf{0}$), we define: 
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \inf \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \inf \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \alpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \frac{1}{\sqrt{k}} \min \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^{|S|} \right\},
\end{align} 
%
where we write $L_{k}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{We note that $1 - \sqrt{k} L_k(\mathbf{M})$ is known as the asymmetric lower restricted isometry constant for matrices $\mathbf{M}$ with unit $\ell_2$-norm columns \cite{Blanchard2011}.} Clearly, $L_{2k}(\mathbf{M}) > 0$ for $\mathbf{M}$ satisfying \eqref{SparkCondition} and $L_{k'}(\mathbf{M}) \geq L_k(\mathbf{M})$ whenever $k' \leq k$. Note also that $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$ if $\mathcal{H} \subseteq {[m] \choose k}$ has $\cup \mathcal{H} = [m]$.

A vector $\mathbf{x}$ is said to be \emph{supported} in $S \subseteq [m]$ when $\mathbf{x} \in \text{\rm span}\{\mathbf{e}_j: j\in S\}$, where $\mathbf{e}_j$ form the standard column basis in $\mathbb R^m$. 
%Denote by $\mathbf{x}^J$ the subvector formed from the entries of $\mathbf{x}$ indexed by $J$. 
A set of $k$-sparse vectors is said to be in \emph{general linear position} when any $k$ of them are linearly independent. The following is a precise statement of our main result.  We leave the quantity $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, \mathcal{H})$ undefined until Eq.~\eqref{Cdef1}.

@@

If $\bar m < mr / (r-1)$ then, letting $p := \bar m - r( \bar m - m)$, we have $|T_1| = \bar m (r - 1) + p$ for positive $p$. By the pigeonhole principle again, there must then be at least $r$ elements of $T_1$ sharing the same first index. By \eqref{cond}, the intersection of the set $\mathcal{G}_1$ consisting of their second indices is nonempty; hence, $|\cap \mathcal{G}_1| = 1$ by $r$-regularity and the SIP. Since $\cap \pi(\mathcal{G}_1)$ is nonempty, applying \eqref{cond} again gives $\cap \pi(\mathcal{G}_1) = \{i_1\}$ for some $i_1 \in [m]$. If $p=1$, then we are done. Otherwise, define $T_2 := T_1 \setminus \{(i,S) \in T_1: i = i_1\}$, which contains $|T_2| = |T_1| - r = (\bar m - 1)(r-1) + (p-1)$ ordered pairs having $\bar m - 1$ distinct first indices. Repeating the above arguments produces a (necessarily) distinct index $i_2$. Iterating this $p$ times yields the set of singletons \mbox{$J = \{i_1, \ldots, i_p\} \subseteq [m]$}.
\end{proof}
